---
title: "Beetle population dynamics: predictors selection (tmp, tmp_z, spei12)"
author: "Maria Potterf"
date: "2024-10-04"
output: 
  html_document:
    toc: true           # Enables the table of contents
    toc_depth: 2        # Sets the level of headers included in the TOC
    toc_float: true        # Makes the TOC float and always visible while scrolling
    number_sections: true  # Enables numbering of sections
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r def-fuctions, echo = F}


# Function to plot a single smooth term
plot_single_smooth <- function(model, variable, x_label = variable, color = "blue") {
  # Generate predictions
  pred <- ggpredict(model, terms = variable)
  
  # Create plot
  p <- ggplot(pred, aes(x = x, y = predicted)) +
    geom_line(color = color) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, fill = color) +
    labs(x = x_label, y = "Predicted") +
    theme_minimal()
  
  return(p)
}


# Function to plot interaction between two smooth terms
plot_interaction_smooth <- function(model, variable1, variable2, x_label = variable1, color = "group", fill = "group") {
  # Generate predictions
  pred <- ggpredict(model, terms = c(variable1, variable2))
  
  # Create plot
  p <- ggplot(pred, aes(x = x, y = predicted)) +
    geom_line( aes(color = !!sym(color))) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = !!sym(fill)), alpha = 0.2) +
    labs(x = x_label, y = "Predicted", color = variable2) +
    theme_minimal()
  
  return(p)
}


```

# Motivation & goal

- when using the previous years values of beetle population, the effect of tmp can be already accounted for. Therefore, should use curent year tmp/tmp_z. what SPEI lag is the better to use?

- tested SPEI12_lag0, SPEI12_lag1,SPEI12_lag2 > SPEI12_lag1 has the most interpretable results

- rerun the models with current year temperature (_lag0) and previous year SPEI12 (_lag1). 

- test if tmp_z or tmp is performing better? in combination 'tmp_z + spei' vs 'tmp + spei'

- create individual models and compare their performance:

Dependent vars: 

- SUM_IPS 
- AGG_DOY 
- PEAK_DOY 
- PEAK_DIFF 


Predictors

- SPEI12 (lag1)
- tmp    (lag0)
- tmp_z  (lag0)


```{r load_models, echo = F}
getwd()
# Load the models and data from the .RData file
load("outData/models_and_data.RData")
```


```{r libs, echo = T}
library(ggplot2)
library(ggeffects)
library(ggpubr)
library(MuMIn)
library(gratia)

```

# Results 

## Find best SPEI LAG

### Summary tmp_z spei12_lag0


```{r summary-lags0}
summary(m.previous.tmp_z0_spei12_0$gam)


```

### Summary tmp_z spei12_lag1
```{r summary-lags1}
summary(m.previous.tmp_z0_spei12_1$gam)

```

### Summary tmp_z spei12_lag2

```{r summary-lags2}
summary(m.previous.tmp_z0_spei12_2$gam)


```

### AIC: compare SPEI lags

```{r get-AIC-population-level-lags}
AIC(m.previous.tmp_z0_spei12_0,
    m.previous.tmp_z0_spei12_1,
    m.previous.tmp_z0_spei12_2)

```

#### Plot SPEI lag 0

```{r test_lags0, echo = F}
m <- m.previous.tmp_z0_spei12_0$gam

# Example usage for a single variable
p_tmp <- plot_single_smooth(m, "tmp_z_lag0", "Temperature Z", color = "blue")
p_spei <- plot_single_smooth(m, "spei12_lag0", "SPEI", color = "red")
p_sum_ips_lag1 <- plot_single_smooth(m, "sum_ips_lag1", "Sum IpS Lag 1", color = "green")

# Example usage for interaction plot
p_interaction <- plot_interaction_smooth(m, "tmp_z_lag0", "spei12_lag0", "Temperature")

# Arrange the plots together
ggarrange(p_tmp, p_spei, p_sum_ips_lag1, p_interaction)
```


#### Plot SPEI lag 1

```{r test_lags1, echo = F}
m <- m.previous.tmp_z0_spei12_1$gam

# Example usage for a single variable
p_tmp <- plot_single_smooth(m, "tmp_z_lag0", "Temperature Z", color = "blue")
p_spei <- plot_single_smooth(m, "spei12_lag1", "SPEI", color = "red")
p_sum_ips_lag1 <- plot_single_smooth(m, "sum_ips_lag1", "Sum IpS Lag 1", color = "green")

# Example usage for interaction plot
p_interaction <- plot_interaction_smooth(m, "tmp_z_lag0", "spei12_lag1", "Temperature Z")

# Arrange the plots together
ggarrange(p_tmp, p_spei, p_sum_ips_lag1, p_interaction)
```

#### Plot SPEI lag 2

```{r test_lags2, echo = F}
m <- m.previous.tmp_z0_spei12_2$gam

# Example usage for a single variable
p_tmp <- plot_single_smooth(m, "tmp_z_lag0", "Temperature Z", color = "blue")
p_spei <- plot_single_smooth(m, "spei12_lag2", "SPEI", color = "red")
p_sum_ips_lag1 <- plot_single_smooth(m, "sum_ips_lag1", "Sum IpS Lag 2", color = "green")

# Example usage for interaction plot
p_interaction <- plot_interaction_smooth(m, "tmp_z_lag0", "spei12_lag2", "Temperature Z")

# Arrange the plots together
ggarrange(p_tmp, p_spei, p_sum_ips_lag1, p_interaction)
```


## Beetle population level

### Pop. level: Summary tmp_z


```{r m.previous.tmp_z0_spei12_1$gam_summary, echo=FALSE}
summary( m.previous.tmp_z0_spei12_1$gam)

``` 

### Pop. level: Summary tmp


```{r m.previous.tmp0_spei12_1$gam_summary, echo=FALSE}
summary( m.previous.tmp0_spei12_1$gam)

``` 

### Pop. level: AIC compare tmp vs tmp_z


```{r get-AIC-population-level}
AIC(m.previous.tmp0_spei12_1,
    m.previous.tmp_z0_spei12_1)

```

### Pop. level: Temporal autocorrelation check

```{r ips_sum_temp_aut_check,  echo = F}
# Extract residuals from the model
residuals <- resid(m.previous.tmp0_spei12_1$lme, type = "normalized")


# Check for temporal autocorrelation
par(mfrow = c(1, 2))  # Arrange plots in one row, two columns
acf(residuals, main = "ACF of Residuals")  # Autocorrelation function
pacf(residuals, main = "PACF of Residuals")  # Partial autocorrelation function

```

### Pop. level: Plot tmp_z 

```{r plot-m.previous.tmp_z0_spei12_1$gam, echo=FALSE}
m<- m.previous.tmp_z0_spei12_1$gam

# Example usage for a single variable
p_tmp <- plot_single_smooth(m, "tmp_z_lag0", "Temperature Z", color = "blue")
p_spei <- plot_single_smooth(m, "spei12_lag1", "SPEI", color = "red")
p_sum_ips_lag1 <- plot_single_smooth(m, "sum_ips_lag1", "sum_ips_lag1", color = "green")

# Example usage for interaction plot
p_interaction <- plot_interaction_smooth(m, "tmp_z_lag0", "spei12_lag1", "Temperature Z")

# Arrange the plots together
ggarrange(p_tmp, p_spei, p_sum_ips_lag1, p_interaction)


```

### Pop. level: Plot tmp 

```{r plot-m.previous.tmp0_spei12_1$gam, echo=FALSE}
m<- m.previous.tmp0_spei12_1$gam
# 

# Example usage for a single variable
p_tmp <- plot_single_smooth(m, "tmp_lag0", "Temperature", color = "blue")
p_spei <- plot_single_smooth(m, "spei12_lag1", "SPEI", color = "red")
p_sum_ips_lag1 <- plot_single_smooth(m, "sum_ips_lag1", "sum_ips_lag1", color = "green")

# Example usage for interaction plot
p_interaction <- plot_interaction_smooth(m, "tmp_lag0", "spei12_lag1", "Temperature")

# Arrange the plots together
ggarrange(p_tmp, p_spei, p_sum_ips_lag1, p_interaction)


```




## Aggregation timing

### Aggregation timing: Summary tmp

```{r summary-tmp-m.agg.previous_tmp0_spei12_1_year_re, echo = T}

summary(m.agg.previous_tmp0_spei12_1$gam)

```
### Aggregation timing: Summary tmp_z

```{r summary-tmp_z-m.agg.previous_tmp0_spei12_1_year_re, echo = T}

summary(m.agg.previous_tmp_z0_spei12_1$gam)

```

### Aggregation timing: AIC compare tmp vs tmp_z

```{r get-AIC_aggregation-timing}
 AIC(#m.agg.previous_tmp0_spei12_1_year_re,   # without random effect is better
  m.agg.previous_tmp0_spei12_1,           # tmp is better
  m.agg.previous_tmp_z0_spei12_1)

```

### Aggregation timing: Temporal autocorrelation check

```{r agg_timing_temp_aut_check, echo = F}
# Extract residuals from the model
residuals <- resid(m.agg.previous_tmp0_spei12_1$lme, type = "normalized")


# Check for temporal autocorrelation
par(mfrow = c(1, 2))  # Arrange plots in one row, two columns
acf(residuals, main = "ACF of Residuals")  # Autocorrelation function
pacf(residuals, main = "PACF of Residuals")  # Partial autocorrelation function

```


### Aggregation timing: Plot aggregation timing


```{r plot-agg-timing, echo = F}

m<- m.agg.previous_tmp0_spei12_1$gam

# Example usage for a single variable
p_tmp <- plot_single_smooth(m, "tmp_lag0", "Temperature", color = "blue")
p_spei <- plot_single_smooth(m, "spei12_lag1", "SPEI", color = "red")
p_sum_ips_lag1 <- plot_single_smooth(m, "tr_agg_doy_lag1", "tr_agg_doy_lag1", color = "green")

# Example usage for interaction plot
p_interaction <- plot_interaction_smooth(m, "tmp_lag0", "spei12_lag1", "Temperature")

# Arrange the plots together
ggarrange(p_tmp, p_spei, p_sum_ips_lag1, p_interaction)

```

## Peak swarming timing

### Peak swarming timing: Summary tmp
```{r summary-tmp-m.peak.doy.previous_tmp0_spei12_1_bam, echo = T}

summary(m.peak.doy.previous_tmp0_spei12_1_bam)

```

### Peak swarming timing: Summary tmp_z

```{r summary-tmp_z-m.peak.doy.previous_tmp0_spei12_1_bam, echo = T}

summary(m.peak.doy.previous_tmp_z0_spei12_1_bam)

```

### Peak swarming timing : AIC Compare tmp vs tmp_z
```{r get-AIC_peak-swarming-timing}
AIC(m.peak.doy.previous_tmp0_spei12_1_bam,   # tmp is better 
  m.peak.doy.previous_tmp_z0_spei12_1_bam)

```

### Peak swarming timing: Temporal autocorrelation check

```{r peak_timing_temp_aut_check, echo = F}
# Extract normalized residuals from the bam model
residuals <- resid(m.peak.doy.previous_tmp0_spei12_1_bam)

# Check for temporal autocorrelation
par(mfrow = c(1, 2))  # Arrange plots in one row, two columns
acf(residuals, main = "ACF of Residuals")  # Autocorrelation function
pacf(residuals, main = "PACF of Residuals")  # Partial autocorrelation function

```


### Peak swarming timing: Plot tmp_z model

```{r plot-tmp_z-peak_swarming_timing, echo = F}

m<-m.peak.doy.previous_tmp_z0_spei12_1_bam

# Example usage for a single variable
p_tmp <- plot_single_smooth(m, "tmp_z_lag0", "Temperature Z", color = "blue")
p_spei <- plot_single_smooth(m, "spei12_lag1", "SPEI", color = "red")
p_sum_ips_lag1 <- plot_single_smooth(m, "tr_peak_doy_lag1", "Peak DDOY Lag 1", color = "green")

# Example usage for interaction plot
p_interaction <- plot_interaction_smooth(m, "tmp_z_lag0", "spei12_lag1", "Temperature Z" )

# Arrange the plots together
ggarrange(p_tmp, p_spei, p_sum_ips_lag1, p_interaction)

```




## Peak swarming intensity

### Peak swarming intensity: Summary tmp

```{r summary-tmp-peak_swarming_intensity}

summary(m.peak.diff.previous.tmp_0_spei12_1$gam)
```

### Peak swarming intensity: tmp_z

```{r summary-tmp_z-peak_swarming_intensity}

summary(m.peak.diff.previous.tmp_z_0_spei12_1$gam)

```

### Peak swarming intensity: AIC Compare 

```{r get-AIC_peak-diff}
 AIC(m.peak.diff.previous.tmp_z_0_spei12_1,
     m.peak.diff.previous.tmp_0_spei12_1)

```


### Peak swarming intensity: Temporal autocorrelation check

```{r peak-diff_temp_aut_check, echo = F}
# Extract residuals from the model
residuals <- resid(m.peak.diff.previous.tmp_0_spei12_1$lme, type = "normalized")


# Check for temporal autocorrelation
par(mfrow = c(1, 2))  # Arrange plots in one row, two columns
acf(residuals, main = "ACF of Residuals")  # Autocorrelation function
pacf(residuals, main = "PACF of Residuals")  # Partial autocorrelation function

```


### Peak swarming intensity: Plot tmp_z model

```{r plot-tmp_z-peak_swarming_intensity, echo = F}

m<-m.peak.diff.previous.tmp_z_0_spei12_1$gam

# Example usage for a single variable
p_tmp <- plot_single_smooth(m, "tmp_z_lag0", "Temperature Z", color = "blue")
p_spei <- plot_single_smooth(m, "spei12_lag1", "SPEI", color = "red")
p_sum_ips_lag1 <- plot_single_smooth(m, "peak_diff_lag1", "Peak Difference Lag 1", color = "green")

# Example usage for interaction plot
p_interaction <- plot_interaction_smooth(m, "tmp_z_lag0", "spei12_lag1", "Temperature")

# Arrange the plots together
ggarrange(p_tmp, p_spei, p_sum_ips_lag1, p_interaction)

```


### Peak swarming intensity: Plot tmp

```{r peak_swarming_intensity_tmp, echo = F}

m <- m.peak.diff.previous.tmp_0_spei12_1$gam

# Example usage for a single variable
p_tmp <- plot_single_smooth(m, "tmp_lag0", "Temperature", color = "blue")
p_spei <- plot_single_smooth(m, "spei12_lag1", "SPEI", color = "red")
p_sum_ips_lag1 <- plot_single_smooth(m, "peak_diff_lag1", "Peak Difference Lag 1", color = "green")

# Example usage for interaction plot
p_interaction <- plot_interaction_smooth(m, "tmp_lag0", "spei12_lag1", "Temperature")

# Arrange the plots together
ggarrange(p_tmp, p_spei, p_sum_ips_lag1, p_interaction)


```

# Conclusion

What SPEI lag to use?

- combination of tmp_lag0 + SPEI_lag1 is the most interpretable and comparable in performance to SPEI_lag2 - use this one

What predictors (tmp vs tmp_z) to use?

- SUM_IPS - tmp_z is better (lower AIC, better R2)
- AGG_DOY - tmp seems better (lower AIC, higher R2)
- PEAK_DOY - tmp is better
- PEAK_DIFF - same AIC, tmp explains a bit more  


| Dependent | AIC: tmp_z | AIC: tmp | R-adj: tmp_z | R-adj: tmp  | best |
|----------|----------|----------|----------|----------|----------|
| Pop. level    | 570  | 584  | 42.7  | 40.8 | tmp_z |
| Aggregation timing    | 407 | 408  |  30.3 | 32.8 | tmp |
| Peak sw. timing    | -631  | -648  | 12.8 | 8.97 | tmp |
| Peak sw intensity    | 704  | 705  | 24.5 | 23.3 | tmp |



```{r table, echo = F}

# Load required packages
library(broom)  # For tidying model output

# Define a function to extract key metrics from a model
# Function to extract R-squared, Adjusted R-squared, AIC, and p-values from a model
extract_model_info <- function(model) {
  # If it's a 'gamm' model, extract the GAM part
  if ("gamm" %in% class(model)) {
    model_summary <- summary(model$gam)  # summary of the GAM part in 'gamm' models
    aic_value <- AIC(model$lme)           # AIC of the mixed model part
  } else {
    model_summary <- summary(model)      # summary for 'bam' or 'gam' models
    aic_value <- AIC(model)              # AIC
  }
  
  # Extract R-squared and adjusted R-squared for 'bam' and 'gam' models
  r_squared <- model_summary$r.sq
  adj_r_squared <- model_summary$adj.r.sq
  
  # Extract p-values for smooth terms
  p_values <- tidy(model)$p.value
  
  # Return the metrics as a list
  return(list(R_squared = r_squared, 
              Adj_R_squared = adj_r_squared, 
              AIC = aic_value, 
              P_values = p_values))
}


# Apply the function to each model and extract information
# Apply the function to each model and extract information
# model_info_list <- list(
#   tmp0_spei12_1 = extract_model_info(m.previous.tmp0_spei12_1),
#   tmp_z0_spei12_1 = extract_model_info(m.previous.tmp_z0_spei12_1),
#   
#   peak_diff_tmp_0_spei12_1 = extract_model_info(m.peak.diff.previous.tmp_0_spei12_1),
#   peak_diff_tmp_z_0_spei12_1 = extract_model_info(m.peak.diff.previous.tmp_z_0_spei12_1),
#   
#   agg_tmp0_spei12_1 = extract_model_info(m.agg.previous_tmp0_spei12_1),
#   agg_tmp_z0_spei12_1 = extract_model_info(m.agg.previous_tmp_z0_spei12_1),
#   
#   peak_doy_tmp0_spei12_1_bam = extract_model_info(m.peak.doy.previous_tmp0_spei12_1_bam),
#   peak_doy_tmp_z0_spei12_1_bam = extract_model_info(m.peak.doy.previous_tmp_z0_spei12_1_bam)
# )
# 

```
