---
title: "Beetle Data LWF Report"
author: "Maria Potterf"
date: "2024-12-13"
output: 
  bookdown::html_document2:
    toc: true           # Adds a table of contents
    toc_depth: 3        # Limits TOC to headings level 1 and 2
    number_sections: true  # Adds numbering to the headings
    toc_float: true     # Makes the TOC float on the side for easy navigation
    fig_caption: true    # Enables figure captions with numbering
link-citations: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



```{r libs, echo=FALSE, include = FALSE}
# Input ------------------------------------

rm(list=ls()) 


source('myPaths.R')

### Libs -----------------------------------------------------------
library(dplyr)
library(sf)
library(stringr)
library(ggplot2)
library(ggpubr)



# Colors libs 
library(RColorBrewer)
library(scales)
library(viridis)


library(tidyr)

library(knitr)



### get data -----------------------------------
path = 'C:/Users/ge45lep/Documents/2022_BarkBeetles_Bavaria/rawData/Fwd__Borkenkäferforschung__Datentransfer'
out_path = 'C:/Users/ge45lep/Documents/2022_BarkBeetles_Bavaria'

# Load RData table
load(paste(path, 'BoMo_2015_2021_Rohdaten.RData', sep = "/"))

high_dens_threshold = 3000


```

# Data Overview

Information about the bark beetle monitoring dataset using pheromone traps. Data represents trap counts per DOY and year. Traps are organized by trap pairs.

**CRS for trap locations**: 

- EPSG:3035 (ETRS89 / LAEA Europe) is a projected coordinate system. 
- Projection Type: Lambert Azimuthal Equal-Area (LAEA).
- Unit: Meters


**Trap catch: Full information (from LWF)**


| Column        | Type            | Description                                                                                                                                                                                                                                   |
|---------------|-----------------|-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| objectid      | Int             | Technical field                                                                                                                                                                                                                             |
| aelf          | Chr             | Abbreviation for the responsible forest office (Amt für Ernährung, Landwirtschaft und Forsten, AELF)                                                                                                                                       |
| kont_dat      | Date (POSIXct)  | Date the trap was emptied. There is not a given day of the week for emptying the trap, therefore the emptying interval may vary.                                                                                                             |
| falsto_name   | Chr             | Name of the trap, consists of monsto_name + trap number (1 or 2)                                                                                                                                                                            |
| art           | Chr             | Species Bark beetle (Ips typographus, Pityogenes chalcographus)                                                                                                                                                                             |
| fangmenge     | Int             | Amount of catched beetles                                                                                                                                                                                                                   |
| einheit       | Chr             | Unit – here only number of individuals (Stück). Small quantities have been counted. Large quantities have been estimated by volume: - Buchdrucker (Ips typographus) 1 ml = 40 individuals - Kupferstecher (Pityogenes chalcographus) 1 ml = 550 individuals |
| koederwechsel | Chr             | Has the bait been changed? Ja – yes; Nein - no                                                                                                                                                                                              |
| monsto_name   | Chr             | Name of the pair of trap locations. One monitoring site consists of two trap sites within two traps per bark beetle species.                                                                                                                |
| representativ | Chr             | Was the catch result representative? There are cases where a catch result is not representative: - The trap was damaged - The bait was removed. Within a radius of 200 metres: - A timber harvesting operation took place - A wood pile was laid - Infestation in the surrounding forest |
| globalid      | Chr             | ID-number of a trap location. It is possible that some trap names have more than one globalid. That is the case if the trap was relocated without renaming the monitoring location (monsto_name).                                             |


```{r, read data, echo = F, warning=F, include =F}
xy_sf <- st_read(paste(out_path, "outSpatial/all_traps_3035.gpkg", sep = '/'))
bav_sf <- st_read(paste(out_path, "outSpatial/bavaria.gpkg", sep = '/'))

# Read elevation per trap
xy_sf_elev_4326 <- st_read(paste(out_path, "outSpatial/traps_with_elevation.gpkg", sep = '/'))


# Reproject bav_sf to the CRS of xy_sf
bav_sf_repr <- st_transform(bav_sf, st_crs(xy_sf))

# simplify: keep only one XY per trap

xy_sf_simple <- xy_sf %>%
  dplyr::select(falsto_name, geom) %>%  # Keep only falsto_name and geometry columns
  group_by(falsto_name) %>%      # Group by falsto_name
  slice(1) %>%                   # Retain only one record per group
  ungroup()                      # Ungroup the data for further operations


# Step 1: Add PairID column
xy_sf_simple <- xy_sf_simple %>%
  mutate(pairID = substr(falsto_name, 1, nchar(falsto_name) - 2))

# Step 2: Calculate average coordinates for each PairID
# Extract coordinates as separate columns
xy_sf_simple <- xy_sf_simple %>%
  mutate(x = st_coordinates(geom)[, 1],  # X coordinate
         y = st_coordinates(geom)[, 2]) # Y coordinate

# Group by PairID and calculate mean coordinates
xy_pair_avg <- xy_sf_simple %>%
  as.data.frame() %>% 
  group_by(pairID) %>%
  summarize(x = mean(x), 
            y = mean(y)) %>%
  ungroup() #%>% 
  #st_as_sf()

# Convert avg_coordinates to sf
xy_pair_sf <- xy_pair_avg %>%
  st_as_sf(coords = c("x", "y"), crs = 3035)



```

# Goal

Are the trap pairs representative one to another?  Eg do we need trap pairs, or a single trap per location?

- Q1: Are they correlated? 
- Q2: What is variation between the two traps (within trap pair)? 
- Q3: Do they cross the >3000 threshold at the same time? 


```{r process_input_data, echo = F}
# Rename the table
# keep all records to check the numbers
dat_full <- Daten_B01

dat <- dat_full

# Convert to date
dat$kont_dat <- as.Date(dat$kont_dat)

# Decompose date into year, month, day, and day of year
# filter only IPS data
dat <- dat %>% 
    dplyr::filter(art == "Buchdrucker") %>%  
  dplyr::mutate(year  = lubridate::year(kont_dat), 
                month = lubridate::month(kont_dat), 
                day   = lubridate::day(kont_dat),
                doy   = lubridate::yday(kont_dat) + 1)

nrow(dat)

# Calculate the number of unique years for each trapID
dat_summary <- dat %>%
  group_by(falsto_name) %>%               # Group by trapID
  summarize(unique_years = n_distinct(year, na.rm = TRUE)) # Count unique years



# keep only representatie = Ja
dat <- dat %>% 
  dplyr::filter(representativ == "Ja")

nrow(dat)
# Calculate the number of unique years for each trapID
dat_summary_repres <- dat %>%
  group_by(falsto_name, monsto_name) %>%               # Group by trapID
  summarize(unique_years = n_distinct(year, na.rm = TRUE)) %>% 
  rename(trapID = falsto_name,
         pairID = monsto_name)# Count unique years

dat_summary_repres_pairID <- dat_summary_repres %>% 
  ungroup() %>% 
  dplyr::select(-trapID) %>% 
  distinct()

# Filter to find trapIDs with exactly 7 unique years
trapIDs_with_7_years <- dat_summary_repres %>%
  dplyr::filter(unique_years == 7)
#table(dat_summary_repres$trapID,dat_summary_repres$unique_years)

# Basic statistics

# Clean trap names and extract trap pair number
dat <- dat %>% 
   mutate(falsto_name = gsub(' ', '_', falsto_name)) %>% 
   mutate(falsto_name = gsub("[^A-Za-z0-9_]", "", falsto_name)) %>% 
    mutate(monsto_name = gsub("[^A-Za-z0-9_]", "", monsto_name)) %>% 
   mutate(trap_pair = as.numeric(str_extract(falsto_name, "[0-9]+")))

# Clean globalid by removing parentheses
dat <- dat %>% 
  mutate(globalid =  gsub("\\{|\\}", "", globalid)) %>% 
  dplyr::filter(year !=2014)


```


Notes about dates & years: 

- 2014 - does not have explicit XY information, trap data are pooled into a single globalid (XY) -> removed from database
- 2015:2021 - only kept
- now all trap pairs locations have 2 traps (eg Zwiesel has A,B,C but only one trap, not a trap pair - rather, a triplet. excluded from this analysis)


# Methods

**Process the data:**

- select only IPS data counts
- remove traps pairs that have trap_pair = 3, or A,B,C indication, instead of 1 or 2
- keep only 'Representative = Ja' (69.000 records compared to full dataset: 73.000)
- get counts per trap 1 and trap 2
- convert data from long to wide format to allow comparison of trap counts on pair levels

*Full data*:

- 302 unique traps (falsto_name)
- 159 unique trap pairs (monsto_name)

*Keeping only representative and remove no-rap pairs design*

- 136 trap pairs




```{r clean_up_data, include = F, warning = F, echo = F}

# Select and rename the desired columns
dat_cleaned <- dat %>%
  dplyr::filter(year !=2014) %>% 
   dplyr::filter(month %in% 4:9) %>%  # filter months during vegetation season: some has record in januay, march ..
  dplyr::select(
    year,
    doy,
    #month,
    fangmenge,
    art,
    falsto_name,
    monsto_name,
    trap_pair
  ) %>%
  dplyr::rename(
    count = fangmenge,
    species = art,
    trapID = falsto_name,
    pairID = monsto_name
  )

# Display the first few rows of the cleaned dataset
#head(dat_cleaned)
#print(summary(dat_cleaned))





```

## Evaluate fit between trap pairs 

- Q1: Spearman correlation
- Q2: Coefficient of variation
- Q3: high density indication: by 3000 beetles/recorded

### Q1: Spearman Correlation coefficient per year

Calculates the correlation between the number of beetles caught (`counts`) per trap pair and year, grouped by `pairID`. We used Spearmann correlation. 

The **Spearman correlation** is a non-parametric measure of the strength and direction of the association between two variables. It assesses how well the relationship between the variables can be described by a monotonic function.

Spearman correlation is calculated based on the ranks of the values, rather than the raw data, making it robust to outliers and suitable for both linear and non-linear relationships.

The correlation coefficient (\(\rho\)) ranges from:

- **+1**: Perfect positive monotonic relationship
- **0**: No monotonic relationship
- **-1**: Perfect negative monotonic relationship


### Q2: Coefficient of variation per year 

The **Coefficient of Variation (CV)** is a statistical measure of the relative variability of a dataset. It is calculated as the ratio of the **standard deviation (SD)** to the **mean**, expressed as a percentage:

\[
\text{CV} = \frac{\text{SD}}{\text{Mean}} \times 100
\]

It shows how much the data (beetle counts per trap) varies in relation to the mean over trap pair. A **low CV** indicates low variability (data tightly clustered around the mean), while a **high CV** indicates high variability (data widely spread out).

### Q3: high density indicators fit

Compare if both traps in trap pair overpasses the >3000 beetles/caught/week at the same time (DOY per year). 

- **over** = both traps have > 3000 beetles (fit)
- **no_fit** = only one trap has > 3000 beetles, the other has less (no fit)
- **less** = both traps have less then 3000 beetles (fit) 

As we have several records per year (1-27), I have calculated how often the three categories occurs from total records times per trap. For map display, I have chosen only **over** and **no_fit** categories, where I have chosen the higher share per record. 


## Identify problematic locations

Order the trap pairs from the 1) least correlated, with the 2) highest coefficiet of variation and 3) with no-fit (3000 threshold).

**Identifying Problematic Trap Pairs**

Identify problematic trap pairs that exhibit inconsistencies in their monitoring performance. Specifically, the focus is on traps with:

- *Low Correlation*: Trap pairs that do not show strong agreement in their catch data, as indicated by low Spearman correlation values.
- *High Variability*: Trap pairs with high Coefficient of Variation (CV), indicating inconsistent or widely dispersed counts.
- *Frequent Misalignment* ("no_fit"): Trap pairs that frequently fail to reach the high-density threshold (>3000 beetles) simultaneously, with one trap exceeding the threshold and the other falling below it.

**Steps to Identify Problematic Traps:**

Aggregate Key Metrics:

Calculate the mean and minimum Spearman correlation (mean_cor, min_cor) for each trap pair across all available records.
Compute the mean and maximum CV (mean_cv, max_cv) to measure variability.
Calculate the percentage of records where the trap pair falls into the "no_fit" category (no_fit_percentage).

**Rank Traps:**

Assign ranks to each trap pair based on average value per trap pair over years:

- Mean correlation (mean_cor): Lower correlation gets a higher rank (more "weird").
- Mean CV (mean_cv): Higher variability gets a higher rank.
- Percentage of "no_fit" occurrences (no_fit_percentage): Higher misalignment gets a higher rank.

Combine these ranks to create an *overall_rank score* that summarizes the degree of "weirdness" for each trap pair.

```{r find_trap_tripplets,echo = F, include = T,warning=F}


# Detect locations with ABC traps, instead of trap pairs
locations_with_sublocations <- dat_cleaned %>%
  mutate(base_name = str_remove(pairID, "[A-Z]$")) %>%  # Remove trailing letter
  group_by(base_name) %>%
  dplyr::filter(n_distinct(pairID) > 1) %>%  # Keep only groups with multiple pairIDs
  ungroup() %>%
  distinct(base_name)  # Get unique base names

```

Identify locations that do not have trap pairs, but have instead A,B,C indication: removed from the dataset:

`r print(locations_with_sublocations)`


```{r ips_correlation, include = F, warning = F, echo = F}

# Exclude rows with triplets locations
dat_cleaned_filtered <- dat_cleaned %>%
  dplyr::filter(!grepl(paste(locations_with_sublocations$base_name, collapse = "|"), pairID))


# Split the data into two datasets by species
ips_data <- dat_cleaned_filtered %>% #dat_cleaned_sum_month %>%
  dplyr::filter(species == "Buchdrucker") %>%  
  ungroup(.)



# Reshape data into wide format with counts from trap_pair 1 and 2
ips_wide <- ips_data %>%
  group_by(pairID, species,  year, doy) %>% # Group by necessary columns
  summarise(count_1 = sum(count[trap_pair == 1], na.rm = TRUE),  # Sum counts for trap_pair 1
            count_2 = sum(count[trap_pair == 2], na.rm = TRUE),  # Sum counts for trap_pair 2
            .groups = "drop") %>% # Ungroup the data after summarising
  mutate(beetle_density = case_when(
    count_1 > high_dens_threshold & count_2 > high_dens_threshold ~ 'over',       # Both counts > high_dens_threshold
    count_1 > high_dens_threshold | count_2 > high_dens_threshold ~ 'no_fit',    # Only one count > high_dens_threshold
    count_1 <= high_dens_threshold & count_2 <= high_dens_threshold ~ 'less'     # Both counts <= high_dens_threshold
  )) #


# Define the full year range
final_year_range <- 2018:2021

# Calculate percentages of each class in `beetle_density`
ips_3k_percentages <- ips_wide %>%
  group_by(pairID, year, beetle_density) %>%  # Group by pairID, year, and beetle_density
  summarize(class_count = n(), .groups = "drop") %>%  # Count occurrences of each class
  group_by(pairID, year) %>%  # Group again by pairID and year
  mutate(class_percentage = class_count / sum(class_count) * 100) %>%  # Calculate percentages
  ungroup()

# keep only higher from teh classes: over and no_fit, remve 'less'
ips_3k_percentages_sub <- ips_3k_percentages %>% 
  dplyr::filter(beetle_density != 'less') %>% 
  group_by(pairID, year) %>%
  slice_max(class_percentage, with_ties = TRUE) %>% # keep rows for equal records
  ungroup() %>% 
  dplyr::select(-class_count)


# calculate beetle metrics;

correlations_ips <- ips_wide %>%
  group_by(pairID, year) %>% # Group by pairID and year
  summarize(
    spearm_cor = cor(count_1, count_2, use = "pairwise.complete.obs", method = "spearman"), # Calculate correlation
    n_records = n(), # Count the number of records in each group
    mean_counts = mean(c(count_1, count_2), na.rm = TRUE),           # Mean of counts
    sd_counts = sd(c(count_1, count_2), na.rm = TRUE),               # Standard deviation of counts
    cv = sd_counts / mean_counts * 100                        # Coefficient of Variation (%)
  ) %>%
   ungroup()

correlations_ips <- correlations_ips %>%
  left_join(ips_3k_percentages_sub)


# merge datao to plot them on teh map
corr_year_xy <- xy_pair_sf %>% 
  inner_join(correlations_ips, by = join_by(pairID))




length(unique(dat_cleaned$trapID))
length(unique(dat_cleaned$pairID))

length(unique(dat_full$falsto_name))
length(unique(dat_full$monsto_name))

length(unique(correlations_ips$pairID))



```

# Results 

```{r scatter_plot, warning=F}
correlations_ips %>% 
  ggplot(aes(x = spearm_cor ,
             y = cv,
             color = factor(year))) +
  geom_point() +
  facet_grid(~year)

```

## Overview

```{r trap_counts}
length(unique(dat$monsto_name))
```




```{r summary_tab, include = T, warning = F}

# Calculate summary statistics
summary_table <- correlations_ips %>%
  group_by(year) %>%
  summarize(
    spearm_cor_mean = mean(spearm_cor, na.rm = TRUE),
    spearm_cor_median = median(spearm_cor, na.rm = TRUE),
    spearm_cor_min = min(spearm_cor, na.rm = TRUE),
    spearm_cor_max = max(spearm_cor, na.rm = TRUE),
    cv_mean = mean(cv, na.rm = TRUE),
    cv_median = median(cv, na.rm = TRUE),
    cv_min = min(cv, na.rm = TRUE),
    cv_max = max(cv, na.rm = TRUE)
  )

# Print the summary table
kable(summary_table, caption = "Summary Statistics for Spearman Correlation and CV Counts by Year")

```
## Distribution of the Spearman correlation 

```{r ips_cor_hist, include = T, warning = F, echo = F, fig.cap = "Histogram showing the distribution of Spearman correlation values (`spearm_cor`) by year. Each facet represents a separate year. The dashed grey line at 0.5 serves for visualization purposes."}

ggplot(data = correlations_ips) +
  geom_histogram(aes(x = spearm_cor), fill = 'steelblue', color = 'black', bin = 30) +  # Histogram with fill based on spearm_cor
  facet_wrap(~ year) +  # Create facets for each year
  geom_vline(xintercept = 0.5, lty = 'dashed', color = 'grey') +
  theme_minimal() +  # Use a minimal theme
  labs(
    title = "Distribution of Spearman Correlation by Year",
    x = "Spearman Correlation",
    y = "Frequency"
  )
```





## Coefficient of variation

The data is highly variable.



```{r ips_cv_hist, include = T, warning = F, echo = F, fig.cap = "Histogram showing the distribution of Coefficient of Variation (CV) counts by year. Each facet represents a separate year, with a dashed grey line at CV = 150% to improve visualization."}
# Create the histogram
ggplot(data = correlations_ips) +
  geom_histogram(aes(x = cv), bins = 30, fill = "steelblue", color = "black") +  # Histogram with 30 bins
  facet_wrap(~ year) +  # Create facets for each year
    geom_vline(xintercept = 100, lty = 'dashed', color = 'grey') +
  theme_minimal() +  # Use a minimal theme
  labs(
    title = "Distribution of CV Counts by Year",
    x = "Coefficient of Variation (CV)",
    y = "Frequency"
  )

```



```{r ips_cv_map, include = TRUE, warning = FALSE, echo = FALSE, fig.cap = "Spatial distribution of the Coefficient of Variation (CV) by year. Points represent traps, with colors scaled to indicate CV values, capped at 150."}
cv_scaling = 150
# Create the plot
ggplot() +
  geom_sf(data = bav_sf_repr, fill = 'grey', color = 'black') +  # Add the bav_sf_repr layer as grey
  geom_sf(data = corr_year_xy, aes(color = cv), size = 2) +  # Plot points with color representing CV
  scale_color_viridis_c(
    option = "plasma",
    name = paste("Coefficient of variation\n(capped at", cv_scaling, ")"),
    limits = c(0, cv_scaling),  # Set limits
    oob = scales::squish  # Values above cv_scaling squish to max cv_scaling
  ) +
  facet_wrap(~ year) +  # Facet the plot by year
  theme_void() +  # Minimal theme for better aesthetics
  labs(
    title = "Coefficient of Variation by Year",
    x = "Longitude",
    y = "Latitude"
  )
```

## High density indication fit

Compare if both traps overpasses the >3000 beetles/cought/week at the same time. 

- over = both traps have > 3000 beetles
- no_fit = only one trap has > 3000 beetles, the other has less
- less = both traps have less then 3000 beetles 

### Share of the fit classes per year 

```{r ips_high_dens_hist, include = T, warning = F, echo = F, fig.cap= "Distribution of CV Counts by Year"}
# Create the histogram
ggplot(data = ips_3k_percentages) +
  geom_bar(aes(x =  beetle_density, fill = beetle_density), bins = 30) +  # Histogram with 30 bins
  facet_grid(~ year) +  # Create facets for each year
    geom_hline(yintercept = 50, lty = 'dashed', color = 'grey') +
  theme_minimal() +  # Use a minimal theme
  labs(
    title = "",
    x = "Coefficient of Variation (CV)",
    y = "Frequency"
  ) +
  theme(legend.position = 'bottom')

```

### Distribution of classes over years 

Spatial distribution of the classes showing indication of the simultaneous increase of high beetle density consitions (**over**: <3000 beetles/ trap catch) and **no-fit**, where one trap reaches the high density threshold but not the other. For visualization purposes, I omitted the **lower** fit (both traps having less then 3000 beetles).

```{r ips_3k_map, include = T, warning = F, echo = F}

# Define a new size column to make points with NA smaller
corr_year_xy <- corr_year_xy %>%
  mutate(point_size = ifelse(is.na(beetle_density), 1, 2))  # Smaller size for NA


# Create the plot
ggplot() +
  geom_sf(data = bav_sf_repr, fill = 'grey', color = 'black') +  # Add the bav_sf_repr layer as grey
  geom_sf(data = corr_year_xy, aes(color = beetle_density, size = point_size)) +  # Adjust point size and color
  scale_size_continuous(range = c(1, 2), guide = "none") +  # Control point size
  scale_color_manual(
    values = c("no_fit" = "blue", "over" = "red", "NA" = "white"),  # Set custom colors
    name = "high density indicator fit"
  ) +
  facet_wrap(~ year) +  # Facet the plot by year
  theme_void() +  # Minimal theme for better aesthetics
  labs(
    title = "high density Indicator Fit by Year",
    x = "Longitude",
    y = "Latitude"
  )





```
## Identify problematic traps

select the traps that are covered in last years: 2018-2021, can be of the higher interest for checking in the field.


```{r identify_problematic_traps}
# Aggregate metrics by trap pair
summary_trap_pairs <- correlations_ips %>%
  mutate(beetle_density = ifelse(is.na(beetle_density), "lower", beetle_density)) %>% 
   dplyr::filter(complete.cases(.)) %>%  # Filter rows with complete cases
  ungroup() %>% 
  group_by(pairID) %>%
  dplyr::summarize(
    mean_cor = mean(spearm_cor, na.rm = TRUE),
    min_cor = min(spearm_cor, na.rm = TRUE),
    max_cv = max(cv, na.rm = TRUE),
    mean_cv = mean(cv, na.rm = TRUE),
     record_years = n(), # Count total number of records for the pairID
    years_no_fit = sum(beetle_density == "no_fit", na.rm = TRUE), # Count years in "no_fit"
    no_fit_percentage = (years_no_fit / record_years) * 100 # Calculate percentage based on total records
  ) %>%
  ungroup() %>% 
  dplyr::select(-record_years)

# Rank trap pairs based on specified criteria
ranked_trap_pairs <- summary_trap_pairs %>%
  mutate(
    cor_rank = rank(mean_cor, ties.method = "min"),           # Rank by mean correlation (ascending)
    cv_rank = rank(-mean_cv, ties.method = "min"),            # Rank by mean CV (descending)
    no_fit_rank = rank(-no_fit_percentage, ties.method = "min") # Rank by no_fit percentage (descending)
  ) %>%
  mutate(
    overall_rank = cor_rank + cv_rank + no_fit_rank           # Combined rank across all criteria
  ) %>%
  arrange(overall_rank) %>%   # Arrange by overall rank
   mutate(order = row_number()) #%>%  # Add order column for "weirdness"
   #full_join(dat_summary_repres_pairID)
# add full traps:


```



```{r plot_problematic}

summary_trap_pairs %>% 
  ggplot(aes(x = mean_cor ,
             y = mean_cv)) +
  geom_point() 

```


```{r plot_problematic_year}

summary_trap_pairs %>% 
  ggplot(aes(x = mean_cor ,
             y = mean_cv)) +
  geom_point() 

```

# Output

At pair level: 

- average XY coordinates: *trap_pairs_avg.gpkg*, approximate locations between trap pair
- table: with full records: *ips_wide_full.csv*
- table on pair records: *correlation_trap_pair.csv*
- table on shares per class: *share_high_density.csv*

**correlation_trap_pair.csv**
 
| Column           | Type   | Description                                                                 |
|-------------------|--------|-----------------------------------------------------------------------------|
| pairID           | Chr    | Identifier for the trap pair location.                                      |
| year             | Num    | Year of data collection.                                                   |
| spearm_cor       | Num    | Spearman correlation coefficient between trap counts.                      |
| n_records     | Int    | Number of trap data collections per year.           |
| mean_counts      | Num    | Mean of beetle counts from both traps in the pair.                         |
| sd_counts        | Num    | Standard deviation of beetle counts from both traps in the pair.           |
| cv       | Num    | Coefficient of Variation (CV), calculated as SD/Mean × 100.                |
| beetle_density          | Chr    | Categorization of counts based on the 3000 threshold: 'over' = both over 3000, 'no_fit' = one over 3000, where the one with the higher sgare is selected. 'lower' is excluded|
| class_percentage | Num    | Percentage of the specific `beetle_density` class within all observations for the year. ('lower' is excluded)|



**share_high_density.csv** 

| Column           | Type   | Description                                                                 |
|-------------------|--------|-----------------------------------------------------------------------------|
| pairID           | Chr    | Identifier for the trap pair location.                                      |
| year             | Num    | Year of data collection.                                                   |
| beetle_density          | Chr    | Categorization of counts based on the 3000 threshold: 'over', 'no_fit', or 'less'. |
| class_count      | Int    | Number of records falling into a specific `beetle_density` category.              |
| class_percentage | Num    | Percentage of records in each `beetle_density` category for a given year and pair. |


**ips_wide_full.csv**

| Column           | Type   | Description                                                                 |
|-------------------|--------|-----------------------------------------------------------------------------|
| pairID           | Chr    | Identifier for the trap pair location.                                      |
| species          | Chr    | Species name (e.g., "Buchdrucker").                                         |
| year             | Num    | Year of data collection.                                                   |
| doy              | Num    | Day of the year (DOY) when the observation was recorded.                   |
| count_1          | Int    | Count of beetles caught by trap 1 in the pair.                             |
| count_2          | Int    | Count of beetles caught by trap 2 in the pair.                             |
| beetle_density          | Chr    | Categorization of counts based on the 3000 threshold: 'over', 'no_fit', or 'less'. |





## Export data
```{r export, echo = F, warning=F}

# Export ips_wide to a CSV file
write.csv(ips_wide, "LWF_report_share/ips_wide_full.csv", row.names = FALSE)

write.csv(correlations_ips, "LWF_report_share/correlation_trap_pair.csv", row.names = FALSE)
# Export ips_wide to a CSV file
write.csv(ips_3k_percentages, "LWF_report_share/share_high_density.csv", row.names = FALSE)



st_write(xy_pair_sf, "LWF_report_share/trap_pairs_avg.gpkg", layer = "layer", 
         driver = "GPKG",  delete_layer = TRUE )



```
