---
title: "Beetle Data LWF Report"
author: "Maria Potterf"
date: "2024-12-13"
output: 
  html_document:
    toc: true           # Adds a table of contents
    toc_depth: 3        # Limits TOC to headings level 1 and 2
    number_sections: true  # Adds numbering to the headings
    toc_float: true     # Makes the TOC float on the side for easy navigation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



```{r libs, echo=FALSE, include = FALSE}
# Input ------------------------------------

rm(list=ls()) 


source('myPaths.R')

### Libs -----------------------------------------------------------
library(lubridate)
library(dplyr)
library(sf)
library(stringr)
library(ggplot2)
library(zoo)
library(data.table)
library(ggpubr)
#library(ggpubr)   # add formulas using  stat_regline_equation(label.x=30000, label.y=40000)
library(ggpmisc)  # add equation to plots smooth 



# Colors libs 
library(RColorBrewer)
library(scales)
library(viridis)


library(tidyr)



### get data -----------------------------------
path = 'C:/Users/ge45lep/Documents/2022_BarkBeetles_Bavaria/rawData/Fwd__Borkenkäferforschung__Datentransfer'
out_path = 'C:/Users/ge45lep/Documents/2022_BarkBeetles_Bavaria'

# Load RData table
load(paste(path, 'BoMo_2015_2021_Rohdaten.RData', sep = "/"))


```

# Data Overview

Information about the bark beetle monitoring  using pheromone traps. 

*CRS*: 

- EPSG:3035 (ETRS89 / LAEA Europe) is a projected coordinate system. 
- Projection Type: Lambert Azimuthal Equal-Area (LAEA).
- Unit: Meters


*Trap catch: Detailed information (from LWF)*


| Column        | Type            | Description                                                                                                                                                                                                                                   |
|---------------|-----------------|-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| objectid      | Int             | Technical field                                                                                                                                                                                                                             |
| aelf          | Chr             | Abbreviation for the responsible forest office (Amt für Ernährung, Landwirtschaft und Forsten, AELF)                                                                                                                                       |
| kont_dat      | Date (POSIXct)  | Date the trap was emptied. There is not a given day of the week for emptying the trap, therefore the emptying interval may vary.                                                                                                             |
| falsto_name   | Chr             | Name of the trap, consists of monsto_name + trap number (1 or 2)                                                                                                                                                                            |
| art           | Chr             | Species Bark beetle (Ips typographus, Pityogenes chalcographus)                                                                                                                                                                             |
| fangmenge     | Int             | Amount of catched beetles                                                                                                                                                                                                                   |
| einheit       | Chr             | Unit – here only number of individuals (Stück). Small quantities have been counted. Large quantities have been estimated by volume: - Buchdrucker (Ips typographus) 1 ml = 40 individuals - Kupferstecher (Pityogenes chalcographus) 1 ml = 550 individuals |
| koederwechsel | Chr             | Has the bait been changed? Ja – yes; Nein - no                                                                                                                                                                                              |
| monsto_name   | Chr             | Name of the pair of trap locations. One monitoring site consists of two trap sites within two traps per bark beetle species.                                                                                                                |
| representativ | Chr             | Was the catch result representative? There are cases where a catch result is not representative: - The trap was damaged - The bait was removed. Within a radius of 200 metres: - A timber harvesting operation took place - A wood pile was laid - Infestation in the surrounding forest |
| globalid      | Chr             | ID-number of a trap location. It is possible that some trap names have more than one globalid. That is the case if the trap was relocated without renaming the monitoring location (monsto_name).                                             |


```{r, read data, echo = F, warning=F, include =F}
xy_sf <- st_read(paste(out_path, "outSpatial/all_traps_3035.gpkg", sep = '/'))
bav_sf <- st_read(paste(out_path, "outSpatial/bavaria.gpkg", sep = '/'))

# Reproject bav_sf to the CRS of xy_sf
bav_sf_repr <- st_transform(bav_sf, st_crs(xy_sf))

#st_crs(xy_sf) == st_crs(bav_sf_repr)
#plot(xy_sf["falsto_name"])

# simplify: keep only one XY per trap

xy_sf_simple <- xy_sf %>%
  dplyr::select(falsto_name, geom) %>%  # Keep only falsto_name and geometry columns
  group_by(falsto_name) %>%      # Group by falsto_name
  slice(1) %>%                   # Retain only one record per group
  ungroup()                      # Ungroup the data for further operations


# Step 1: Add PairID column
xy_sf_simple <- xy_sf_simple %>%
  mutate(pairID = substr(falsto_name, 1, nchar(falsto_name) - 2))

# Step 2: Calculate average coordinates for each PairID
# Extract coordinates as separate columns
xy_sf_simple <- xy_sf_simple %>%
  mutate(x = st_coordinates(geom)[, 1],  # X coordinate
         y = st_coordinates(geom)[, 2]) # Y coordinate

# Group by PairID and calculate mean coordinates
xy_pair_avg <- xy_sf_simple %>%
  as.data.frame() %>% 
  group_by(pairID) %>%
  summarize(x = mean(x), 
            y = mean(y)) %>%
  ungroup() #%>% 
  #st_as_sf()

# Convert avg_coordinates to sf
xy_pair_sf <- xy_pair_avg %>%
  st_as_sf(coords = c("x", "y"), crs = 3035)



```

# Goal:

Are the trap pairs representative one to another?  Eg do we need trap pairs, or a single trap per location?

- are they correlated?
- what is difference between the two traps (within trap pair)?
- do they cross the >3000 threshold at the same time? 


```{r process_input_data, echo = F}
# Rename the table
dat <- Daten_B01

# Convert to date
dat$kont_dat <- as.Date(dat$kont_dat)

# Decompose date into year, month, day, and day of year
dat <- dat %>% 
  dplyr::mutate(year  = lubridate::year(kont_dat), 
                month = lubridate::month(kont_dat), 
                day   = lubridate::day(kont_dat),
                doy   = lubridate::yday(kont_dat) + 1)

# keep only representatie = Ja
dat <- dat %>% 
  dplyr::filter(representativ == "Ja")

# Basic statistics
#nrow(dat) # Total rows >73,000
#length(unique(dat$falsto_name)) # Total traps = 302

# Clean trap names and extract trap pair number
dat <- dat %>% 
   mutate(falsto_name = gsub(' ', '_', falsto_name)) %>% 
   mutate(falsto_name = gsub("[^A-Za-z0-9_]", "", falsto_name)) %>% 
    mutate(monsto_name = gsub("[^A-Za-z0-9_]", "", monsto_name)) %>% 
   mutate(trap_pair = as.numeric(str_extract(falsto_name, "[0-9]+")))

# Clean globalid by removing parentheses
dat <- dat %>% 
  mutate(globalid =  gsub("\\{|\\}", "", globalid)) %>% 
  dplyr::filter(year !=2014)
```


Recording dates & years: 

- 2014 - does not have explicit XY information, trap data are pooled into a single globalid (XY)
- 2015:2021


In total, 138 traps has changed location 2-4 times over 2015-2021 from total of 302 locations

# Methods

**Process the data:**

- select only IPS data counts
- remove traps pairs that have trap_pair = 3, or A,B,C indication, instead of 1 or 2
- keep only 'Representative = Ja' (69.000 records compared to full dataset: 73.000)
- get counts per trap 1 and trap 2
- convert data from long to wide format to allow to compare trap counts on pair levels


```{r clean_up_data, include = F, warning = F, echo = F}

# Select and rename the desired columns
dat_cleaned <- dat %>%
  dplyr::filter(year !=2014) %>% 
   dplyr::filter(month %in% 4:9) %>%  # filter months during vegetation season: some has record in januay, march ..
  dplyr::select(
    year,
    doy,
    #month,
    fangmenge,
    art,
    falsto_name,
    monsto_name,
    trap_pair
  ) %>%
  dplyr::rename(
    count = fangmenge,
    species = art,
    trapID = falsto_name,
    pairID = monsto_name
  )

# Display the first few rows of the cleaned dataset
#head(dat_cleaned)
print(summary(dat_cleaned))


```

## Metrics 

- Spearman correlation
- Coefficient of variation
- Outbreak indication

### Spearman Correlation coefficient per year

Calculates the correlation between the number of beetles caught (`counts`) per trap pair and year, grouped by `pairID`. We used Spearmann correlation. 

The **Spearman correlation** is a non-parametric measure of the strength and direction of the association between two variables. It assesses how well the relationship between the variables can be described by a monotonic function.

Spearman correlation is calculated based on the ranks of the values, rather than the raw data, making it robust to outliers and suitable for both linear and non-linear relationships.

The correlation coefficient (\(\rho\)) ranges from:
- **+1**: Perfect positive monotonic relationship
- **0**: No monotonic relationship
- **-1**: Perfect negative monotonic relationship


### Coefficient of variation per year 

The **Coefficient of Variation (CV)** is a statistical measure of the relative variability of a dataset. It is calculated as the ratio of the **standard deviation (SD)** to the **mean**, expressed as a percentage:

\[
\text{CV} = \frac{\text{SD}}{\text{Mean}} \times 100
\]

It shows how much the data (beetle counts per trap) varies in relation to the mean over trap pair. A **low CV** indicates low variability (data tightly clustered around the mean), while a **high CV** indicates high variability (data widely spread out).

### Outbreak indicators fit

Compare if both traps in trap pair overpasses the >3000 beetles/caught/week at the same time (DOY per year). 

- over = both traps have > 3000 beetles
- no_fit = only one trap has > 3000 beetles, the other has less
- less = both traps have less then 3000 beetles 

As we have several records per year (1-27), I have calculated how often the three categories occurs from total records times per trap. For map display, I have chosen only 'over' and 'no_fit' categories, where I have chosen the higher share per record. 




```{r ips_correlation, include = F, warning = F, echo = F}

# Split the data into two datasets by species
ips_data <- dat_cleaned %>% #dat_cleaned_sum_month %>%
  dplyr::filter(species == "Buchdrucker") %>% 
  ungroup(.)

# Reshape data into wide format with counts from trap_pair 1 and 2
ips_wide <- ips_data %>%
  group_by(pairID, species, year, doy) %>% # Group by necessary columns
  summarise(count_1 = sum(count[trap_pair == 1], na.rm = TRUE),  # Sum counts for trap_pair 1
            count_2 = sum(count[trap_pair == 2], na.rm = TRUE),  # Sum counts for trap_pair 2
            #count_3 = sum(count[trap_pair == 3], na.rm = TRUE),  # Sum counts for trap_pair 3 (if applicable)
            .groups = "drop") %>% # Ungroup the data after summarising
  mutate(over_3k = case_when(
    count_1 > 3000 & count_2 > 3000 ~ 'over',       # Both counts > 3000
    count_1 > 3000 | count_2 > 3000 ~ 'no_fit',    # Only one count > 3000
    count_1 <= 3000 & count_2 <= 3000 ~ 'less'     # Both counts <= 3000
  ))


# Calculate percentages of each class in `over_3k`
ips_3k_percentages <- ips_wide %>%
  group_by(pairID, year, over_3k) %>%  # Group by pairID, year, and over_3k
  summarize(class_count = n(), .groups = "drop") %>%  # Count occurrences of each class
  group_by(pairID, year) %>%  # Group again by pairID and year
  mutate(class_percentage = class_count / sum(class_count) * 100) %>%  # Calculate percentages
  ungroup()

# keep only higher from teh classes: over and no_fit, remve 'less'
ips_3k_percentages_sub <- ips_3k_percentages %>% 
  dplyr::filter(over_3k != 'less') %>% 
  group_by(pairID, year) %>%
  slice_max(class_percentage, with_ties = TRUE) %>%
  ungroup() %>% 
  dplyr::select(-class_count)



correlations_ips <- ips_wide %>%
  group_by(pairID, year) %>% # Group by pairID and year
  summarize(
    spearm_cor = cor(count_1, count_2, use = "pairwise.complete.obs", method = "spearman"), # Calculate correlation
    record_count = n(), # Count the number of records in each group
    mean_counts = mean(c(count_1, count_2), na.rm = TRUE),           # Mean of counts
    sd_counts = sd(c(count_1, count_2), na.rm = TRUE),               # Standard deviation of counts
    cv_counts = sd_counts / mean_counts * 100                        # Coefficient of Variation (%)
  ) %>%
 
  ungroup()

correlations_ips <- correlations_ips %>%
  left_join(ips_3k_percentages_sub)

# add to original table
# correlations_ips <- correlations_ips %>%
#   left_join(ips_class_percentages %>% 
#               select(pairID, year, over_3k, class_percentage),
#             by = c("pairID", "year"))

# merge datao to plot them on teh map
corr_year_xy <- xy_pair_sf %>% 
  inner_join(correlations_ips, by = join_by(pairID))


```

# Results 

## Overview

```{r summary_tab, include = T, warning = F}

summary(correlations_ips)

```
## Distribution of the Spearman correlation 

```{r ips_cor_hist, include = T, warning = F, echo = F}
ggplot(data = correlations_ips) +
  geom_histogram(aes(x = spearm_cor), fill = 'steelblue', color = 'black', bin = 30) +  # Histogram with fill based on spearm_cor
  facet_wrap(~ year) +  # Create facets for each year
  geom_vline(xintercept = 0.5, lty = 'dashed', color = 'grey') +
  theme_minimal() +  # Use a minimal theme
  labs(
    title = "Distribution of Spearman Correlation by Year",
    x = "Spearman Correlation",
    y = "Frequency"
  )
```





## Coefficient of variation

A CV of **200%** means that the **standard deviation is twice as large as the mean**. In practical terms:
- The data is highly dispersed and inconsistent.
- On average, the values deviate from the mean by **200% of the mean**.

For example, if the mean is 50, a CV of 200% suggests that the data values typically vary by ±100 around the mean, indicating extreme variability.

```{r ips_cv_hist, include = T, warning = F, echo = F}
# Create the histogram
ggplot(data = correlations_ips) +
  geom_histogram(aes(x = cv_counts), bins = 30, fill = "steelblue", color = "black") +  # Histogram with 30 bins
  facet_wrap(~ year) +  # Create facets for each year
    geom_vline(xintercept = 100, lty = 'dashed', color = 'grey') +
  theme_minimal() +  # Use a minimal theme
  labs(
    title = "Distribution of CV Counts by Year",
    x = "Coefficient of Variation (CV)",
    y = "Frequency"
  )

```



```{r ips_cv_map, include = T, warning = F, echo = F}

# Create the plot
ggplot() +
  geom_sf(data = bav_sf_repr, fill = 'grey', color = 'black') +  # Add the bav_sf_repr layer as grey
  geom_sf(data = corr_year_xy, aes(color = cv_counts), size = 2) +  # Plot points with color representing Spearman correlationSpearman correlation
  #scale_color_viridis_c(option = "plasma", name = "Coefficient of variation") +  # Use a continuous color scale
  scale_color_viridis_c(
    option = "plasma",
    name = "Coefficient of variation\n(capped at 150)",
    limits = c(0, 150),  # Set limits
    oob = scales::squish  # Values above 150 squish to max (150)
  ) +
   facet_wrap(~ year) +  # Facet the plot by year
  
  theme_void() +  # Minimal theme for better aesthetics
  labs(
    title = "Coefficient of variation by Year",
    x = "Longitude",
    y = "Latitude"
  )

```

## Outbreak indication fit

Compare if both traps overpasses the >3000 beetles/cught/week at the same time. 

- over = both traps have > 3000 beetles
- no_fit = only one trap has > 3000 beetles, the other has less
- less = both traps have less then 3000 beetles 

### Distribution of the Outbreak indicator fit

```{r ips_outbreak_hist, include = T, warning = F, echo = F}
# Create the histogram
ggplot(data = ips_3k_percentages) +
  geom_bar(aes(x =  over_3k, fill = over_3k), bins = 30) +  # Histogram with 30 bins
  facet_grid(~ year) +  # Create facets for each year
    geom_hline(yintercept = 50, lty = 'dashed', color = 'grey') +
  theme_minimal() +  # Use a minimal theme
  labs(
    title = "Distribution of CV Counts by Year",
    x = "Coefficient of Variation (CV)",
    y = "Frequency"
  ) +
  theme(legend.position = 'bottom')

```


```{r ips_3k_map, include = T, warning = F, echo = F}

# Create the plot
ggplot() +
  geom_sf(data = bav_sf_repr, fill = 'grey', color = 'black') +  # Add the bav_sf_repr layer as grey
  geom_sf(data = corr_year_xy, aes(color = over_3k ), size = 2) +  # Plot points with color representing Spearman correlationSpearman correlation
  scale_color_viridis_d(option = "plasma", 
                        name = "Outbreak indicator fit",
                        na.value = "white") +  # Use a discrete color scale
   facet_wrap(~ year) +  # Facet the plot by year
  theme_void() +  # Minimal theme for better aesthetics
  labs(
    title = "Oubreak indicator fit by Year",
    x = "Longitude",
    y = "Latitude"
  )

```


# Output

at Pair level: 

- average XY coordinates
- table: with full records: 
- table on pair records:




