---
title: "Ips population dynamic in Bavaria (2015-2021)"
author: "Mária Potterf"
date:  "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 2
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../outReports",
      output_file = file.path("../outReports", glue::glue('Beetles_dyn_{Sys.Date()}'
    )))
  })

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, tidy.opts = list(comment = FALSE), message = F, warning = F)
knitr::opts_chunk$set(fig.width=6, fig.height=3) 

```


```{css, echo=FALSE}
p.caption {
  font-size: 0.8em;
}
```


```{r setup-root}

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file('C:/Users/ge45lep/Documents/2022_BarkBeetles_Bavaria/beetlesbavaria'))

```


```{r, check-dir, include=FALSE}
#knitr::opts_knit$set(root.dir = "C:/Users/ge45lep/Documents/2021_Franconia_mortality/r_franconia_mortality")

#getwd()
```


```{r, read-data, include = F}

load(file = "outData/ips.Rdata")
load(file = "outData/spatial.Rdata")
```

# To explore

Variation of beetle population dynamics over time and space: how does the beetle counts change over time and space?

- Ips typographus, maybe Pityogenes chalcographus (?)
- earlier/steeper beetle counts per year?

What are the drivers?

- spruce%
- SPEI (drougt) index
- elevation
- temp
- precipitation

Regional vs global variation?

- are the traps near by more similar than the traps further away?


# Input

## Beetle trap dataset overview

- collected whole year, roughtly in two weeks interval (range 7 - 30 days)
- individual trap ID - same 'globalid' for Ips and Pityogenes

[tbd]

## Counts processing

- count per year = sum of beetles per trap/year over vegetation period
- counts per day =  corrected by the frequency of trap emptying to have average beetles counts per day.


### Filtering: 

- vegetation period: April 1st to Oct 30
- remove:

 + if revisit times is less then 10 per season
 + zero beetles cought per whole season 
 + trap number 3 (not a pair: 1 or 2)
 + only 'representative' == "Ja"
 

# Descriptive stats

```{r}
hist(df.daily$avg_day_count)
```


DOY overview

DOY

- 100 April 10
- 150 May 30
- 200 July 19
- 211 July 30
- 250 SEpt 7

# Temporal variation of beetles counts/traps/day

RQ: Does the increase in population happens earlier in year?

- what is the DOY of the max increase of beetles daily counts between two consecutive days?
- BY how many beetles?





