---
title: "Ips population dynamic in Bavaria (2015-2021)"
author: "Mária Potterf"
date:  "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 2
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../outReports",
      output_file = file.path("../outReports", glue::glue('Beetles_dyn_{Sys.Date()}'
    )))
  })

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, tidy.opts = list(comment = FALSE), message = F, warning = F)
knitr::opts_chunk$set(fig.width=6, fig.height=3) 

```


```{css, echo=FALSE}
p.caption {
  font-size: 0.8em;
}
```


```{r setup-root}

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file('C:/Users/ge45lep/Documents/2022_BarkBeetles_Bavaria/beetlesbavaria'))

```


```{r, check-dir, include=FALSE}
#knitr::opts_knit$set(root.dir = "C:/Users/ge45lep/Documents/2021_Franconia_mortality/r_franconia_mortality")

#getwd()
```


```{r, read-data, include = F}

load(file = "outData/ips.Rdata")
load(file = "outData/spatial.Rdata")
```

# Research questions

Variation of beetle population dynamics over time and space: 

- how does the beetle counts change over time and space?
- does beetle population increase earlier in the year? 
- is the increase steeper over years? 

Regional vs global variation?

- are the traps near by more similar than the traps further away?


# Input data

## Beetle trap dataset overview

- trap datasets in trap pairs, covering area of Bavaria over 2015-2021
- consistent monitoring: 79*2 traps per whole research period
- traps emptied over whole year, roughtly in two weeks interval (range 7 - 30 days)
- individual trap name = 'falsto_name' - one trap can have several globalid, if moved over years. Keep only one globalid per trap. Only one 'globalid' for Ips and Pityogenes traps locations.

## Dependent variables

- ips_sum - sum of beetles per trap/year over vegetation period
- counts per day =  corrected by the frequency of trap emptying to have average beetles counts per day
- DOY of the max population increase
- Max population increase per year/trap
- output of LISA?

## Predictors?

- temp
- precipitation
- SPEI (drought) index
- soil water content (swvl)
- elevation
- spruce%
- spruce vs. deciduous%

### Curate the data: 

- vegetation period: April 1st to Oct 30
- remove:

 + if revisit times is less then 5 per season
 + zero beetles cough per whole season 
 + trap number 3 (not a pair: 1 or 2)


## Descriptive stats

```{r}
hist(df.daily$avg_day_count)
```


DOY overview

DOY

- 100 April 10
- 150 May 30
- 200 July 19
- 211 July 30
- 250 SEpt 7

# Results 

## Temporal variation of beetles counts within a year

RQ: Does the increase in population happens earlier in year?

Calculate the differences in beetle counts between two consecutive dates. Fin the max increase per trap & year. Identify the DOY when this max increase happend.


```{r p_count_diff, fig.width = 5, fig.height = 5, echo = F, fig.cap="Average differences in consecutive beetle counts per year (DOY - day of the year) over the study period (2015:2021)"}

p_count_diff  

```

Scatter plot: DOY vs max increase in beetle numbers per year per trap. One point = one trap

```{r p_diff_doy, fig.width = 4, fig.height = 5, echo = F, fig.cap="Scatter plot of the max increase of beetle count and respective doy over study period."}

p_diff_doy  

```

### Map: Max increase in beetle population per DOY

```{r map_p_doy_max_increase, fig.width = 6, fig.height = 6, echo = F, fig.cap="DOY of max increase of betle counts between two consecutive days and size of beetles numbers increase"}

p_doy_max_increase  

```

### Map: Max increase in beetle population per DOY (DOY April 1st to May 30, DOY 92-150)
```{r map_p_doy_max_increase160, fig.width = 6, fig.height = 6, echo = F, fig.cap="DOY of max increase of betle counts between two consecutive days and size of beetles numbers increase: filtered for DOY < 150 (May 30)"}

p_doy_max_increase150  

```





## Beetle sum counts over years (2015:2021)

### Clim predictors

- temp
- precipitation
- SPEI (drought) index
- soil water content (swvl)
- elevation
- spruce%
- spruce vs. deciduous%

## LISA: measure of similarity between trap and its surrounding

## Global Moran's I - overall measure of spatial autocorrelation over Bavaria

## Semivariance



