---
title: "Ips population dynamic in Bavaria (2015-2021)"
author: "Mária Potterf"
date:  "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 3
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../outReports",
      output_file = file.path("../outReports", glue::glue('Beetles_dyn_{Sys.Date()}'
    )))
  })

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, tidy.opts = list(comment = FALSE), message = F, warning = F)
knitr::opts_chunk$set(fig.width=6, fig.height=3) 

```


```{css, echo=FALSE}
p.caption {
  font-size: 0.8em;
}
```


```{r setup-root}

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file('C:/Users/ge45lep/Documents/2022_BarkBeetles_Bavaria/beetlesbavaria'))

```


```{r, check-dir, include=FALSE}
#knitr::opts_knit$set(root.dir = "C:/Users/ge45lep/Documents/2021_Franconia_mortality/r_franconia_mortality")

#getwd()
```


```{r, read-data, include = F}

load(file = "outData/ips.Rdata")
load(file = "outData/spatial.Rdata")
load(file = "outData/predict.Rdata")
load(file = "outData/lisa.Rdata")


library(dplyr)
library(cowplot)
library(ggplot2)
library(ggpubr)
library(ggpmisc)
library(kableExtra)

```

# Research questions

Q1: Variation of beetle population dynamics over time and space: 

 - Do beetle counts vary between trap pairs? 
 - How do beetle counts change over time and space?
   - Sum beetle counts per trap/year
 - Does beetle population increase earlier in the year? When does the main increase in population (*swarming*) happen?
   - Identify day of the year (DOY) of the maximal difference in beetle counts between two consecutive dates.
 - Do beetles aggregate faster? When they reach the same threshold? (1000 beetles) *aggregation*
   - Identify (DOY) of reaching the 1000 beetles per trap.
   
Q2: Regional vs global variation? - run on 'sum of beetles/year' variable (posible to run on other dependent variables as well)

- are the traps near by more similar than the traps further away?
- LISA - local indicators of spatial association
- Global Moran's I - global indicators of spatial autocorrelation
- variograms - identifies the distance at what two random location are equally different (their difference does not further increase with increasing distance)

Q3: Drivers? (data collected, but not analyzed yet)

- temperature
- precipitation
- soil water content (swvl) - ratio volume water/volume soil: higher the number, more water, less drought
- elevation
- spruce%



# Methods

## Input data

### Beetle trap dataset overview

- trap datasets in trap pairs, covering area of Bavaria over 2015-2021
- cca 10 x 10 km span, 1 km distance one trap from the other in one location
- consistent monitoring: 79*2= 158 traps/year per whole research period
- removed traps without pair (e.g. '3'), also traps that were added later in the study period (mostly in Bavarian forests, north-west Bavaria)
- traps emptied over whole year, roughtly in two weeks interval (range 7 - 30 days)
- filtered trap data from April 1st to Oct 30th
- individual trap name = 'falsto_name' - one trap can have several globalid, if moved over years. Keep only one globalid per trap. Only one 'globalid' for traps locations.





### Dependent variables

- total sum of beetles per trap/year
- beetle **emergence**: the max difference in beetle counts between two consecutive dates 
- beetle **aggregation** : DOY of the beetles reaching 1000 beetles/trap.






```{r summary-tab, caption = 'Summary table f the beetle counts per year and predictors '}
ips_sum_preds %>% 
  dplyr::select(-c(spei1, spei3, spei6, spei12, tpi, tri )) %>% 
summary()

```


```{r p_clim_data, fig.width = 6, fig.height = 3, fig.cap= "Red line indicates local mean."}

ggarrange(p_temp, p_prec)

#cowplot::plot_grid(p_temp, p_prec)
```

## Data processing

### Curate the data 

- vegetation period: April 1st to Oct 30 (DOY start `r doy.start` - DOY end `r doy.end`)
- removed:

  + if revisit times is less then 5 per season
  + zero beetles cough per whole season 
  + trap number 3 (not a pair: 1 or 2)
  + only kept consistent traps, they were surveyed per whole period 2015-2021 -> 79*2 traps in total (158)



### Identify multicollinearity between predictors

VIF - variance inflation factor
values for each predictor variable in the model to determine if multicollinearity

Based on VIF values:

 - less 1  - relatively independent
 - 1-5  - very little dependent
 - more 5  - likely concerning - removed

Based on VIF values, I have removed the most correlated predictors.

Set of the least correlated predictors, which can be further used for analysis:

```{r p_vif_fin, fig.width = 3, fig.height = 3.5, echo = F, fig.cap="Values of the selected predictors based on the low (<5) values of Variance Inflation Factors. Vertical line represent value 1, indicating non-correlated precictors."}

vif_values_fin

#create horizontal bar chart to display each VIF value
barplot(vif_values_fin, main = "VIF Values", horiz = TRUE, col = "steelblue")

#add vertical line at 5
abline(v = 1, lwd = 3, lty = 2)


```

### Check correlation for selected predictores 



```{r lm_temp_elev, fig.width = 5, fig.height = 5, echo = F, fig.cap="Correlation between known predictors: elevation vs temperature"}

df_predictors_year %>% 
  ggplot(aes(x = temp,
             y = elev,
             group = year)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c( "R2", 'p'))) + #"eq",
  geom_point(alpha = 0.5) +
  facet_wrap(.~year) + #, scales = 'free'
  theme_bw()

```


```{r lm_prec_elev, fig.width = 5, fig.height = 5, echo = F, fig.cap="Correlation between known predictors: precipitation vs elevation" }

df_predictors_year %>% 
  ggplot(aes(x = prec,
             y = elev,
             group = year)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c( "R2", 'p'))) + #"eq",
  geom_point(alpha = 0.5) +
  facet_wrap(.~year) + #, scales = 'free'
  theme_bw()

```


```{r lm_prec_freq, fig.width = 5, fig.height = 5, echo = F, fig.cap="Correlation between known predictors: precipitation vs spruce share" }

# precipitation vs spruce share
df_predictors_year %>% 
  ggplot(aes(x = prec,
             y = freq,  # share of spruce
             group = year)) +
  stat_poly_line() +
  #stat_poly_eq(use_label(c( "R2", 'p'))) + #"eq",
  geom_point(alpha = 0.5) +
  facet_wrap(.~year) + #, scales = 'free'
  theme_bw()

```

```{r lm_temp_freq, fig.width = 5, fig.height = 5, echo = F, fig.cap="Correlation between known predictors: temperature vs spruce share (freq)" }

# temp vs spruce share
df_predictors_year %>% 
  ggplot(aes(x = temp,
             y = freq,  # share of spruce
             group = year)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c( "R2", 'p'))) + #"eq",
  geom_point(alpha = 0.5) +
  facet_wrap(.~year) + #, scales = 'free'
  theme_bw()



```


# Results 

DOY overview

   DOY    Date

 - 100 - April 10
 - 150 - May 30
 - 200 - July 19
 - 211 - July 30
 - 250 - Sept 7


## Descriptive stats

### Correlation between trap pairs

Linear regression between traps 1&2.

```{r p_lm_traps, fig.width = 5, fig.height = 5, echo = F, fig.cap="LM of the beetle sums per year between trap 1 and trap 2 (same location)"}
p_lm_traps
```




## Beetle dynamics over time and space

### Beetle counts, emergence and aggregation DOY

```{r p_bars, fig.width = 6, fig.height = 6, echo = F, fig.cap="Beetle population dynamics. (a) total sum per year. (b) DOY of beetle aggregation per trap. (c) DOY of max increase between dates, (d) Max increase in beetle numbers. Shown are medians and IQR. "}


ggarrange(p_ips.year.sum,
          p_ips.agg,
          p_ips.max.diff.doy,
          p_ips.max.diff, labels = 'auto',
          ncol = 2, nrow = 2 )

```


### Temporal variation of beetles counts within a year

RQ: Does the increase in population happens earlier in year?

Calculate the differences in beetle counts between two consecutive dates. Find the max increase per trap & year. Identify the DOY when this max increase happend.


```{r p_count_diff, fig.width = 5, fig.height = 5, echo = F, fig.cap="Average differences in consecutive beetle counts per year (DOY - day of the year) over the study period (2015:2021)"}

#p_count_diff  

```

#### Map: Timing of the beetle swarming  

Max increase in beetle population per DOY (all)

```{r map_p_doy_max_increase, fig.width = 6, fig.height = 6, echo = F, fig.cap="DOY of max increase of betle counts between two consecutive days and size of beetles numbers increase"}

p_doy_max_increase  

```

#### Map: Map: Timing of the beetle swarming (filtered, <150)

Max increase in beetle population per DOY (filtered, <150)
```{r map_p_doy_max_increase160, fig.width = 6, fig.height = 6, echo = F, fig.cap="DOY of max increase of betle counts between two consecutive days and size of beetles numbers increase: filtered for DOY < 150 (May 30)"}

p_doy_max_increase150  

```




#### Summary table: Max increase in beetle population per DOY & max difference
```{r}
max.diff.doy %>% 
  as.data.frame() %>% 
  group_by(year) %>% 
  dplyr::summarise(min    = min(doy),
            max    = max(doy),
            mean   = mean(doy),
            sd     = sd(doy),
            #cv     = sd/mean,
            median = median(doy)) %>% 
  round() %>% 
  knitr::kable("markdown", caption = 'DOY of the max increase in beetle numbers.') %>% 
  #kable("html") %>%
  kable_styling(font_size = 7)


# Check Max increase population
max.diff.doy %>% 
  as.data.frame() %>% 
  group_by(year) %>% 
  dplyr::summarise(min = min(diff),
            max = max(diff),
            mean = mean(diff),
            sd = sd(diff),
           # cv = sd/mean,
            median = median(diff)
            )  %>% 
  round() %>% 
  knitr::kable("markdown", caption = 'Max increase of beetle numbers ')

```




#### Map: Beetle aggregation per DOY (DOY < 130)

```{r map_p_aggreg, fig.width = 6, fig.height = 6, echo = F, fig.cap="Timing of reaching 1000 beetles per trap & year. Selected are DOY <130 for improved visibility."}

p_aggreg  

```



```{r}
ips.aggreg %>% 
  as.data.frame() %>% 
  group_by(year) %>% 
  dplyr::summarise(min    = min(doy),
                   max    = max(doy),
                   mean   = mean(doy),
                   sd     = sd(doy),
                  # cv     = sd/mean,
                   median = median(doy))  %>% 
  round() %>% 
  knitr::kable("markdown", caption = 'DOY of Beetle aggregation (1000)') 
  
```





## Local vs regional variation

### LISA: measure of similarity between trap and its surrounding

10 nearest distance neighbors
LISA identifies the (di)similarity in beetles counts/trap between neighbouring traps 

Categories: 

- High-High - hotspot, having relatively high value, cluster
- High-Low - dissimilar values from neighours
- Low-High - dissimilar values from neighours 
- Low-Low - coldspots, similar to neighbors

Individual years were sliced, so he yearly hotspots vs coldspots are classified within the respective year.


```{r p_lisa_sub, fig.width = 6, fig.height = 6, echo = F, fig.cap="Local Indicators of Spatial Assotiation. Shown are only High-High and Low-Low clusters. No significat values are not shown."}

p_lisa_sub  

```

How many traps are similar vs dissimilar over years?

- share of 'high-high' increases over years
- lowest dissimilarity between traps: 2020, highest in 2016

```{r p_lisa_freq, fig.width = 3, fig.height = 2.5, echo = F, fig.cap="Local Indicators of Spatial Assotiation. Frequency of the similarity categories per trap: High-High, High-Low, Low-High, Low-Low"}

p_lisa_freq +
  theme_bw()

```

### Global Moran's I - overall measure of spatial autocorrelation over Bavaria

```{r p_glob_moran_bar, fig.width = 2.5, fig.height = 2.5, echo = F, fig.cap="Global Moran's I per each year. Higher value indicates higher autocorrelation"}

p_glob_moran_bar  

```

### Semivariance
```{r semi_out, fig.width = 6, fig.height = 6, echo = F, fig.cap="Variograms per year. Note the different variogram fitting (line)"}

semi_out  

```
    



